Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-24 icfp.tex
Start 3.6.2

\subsubsection{Pattern matching on non-WHNF terms via proof irrelevant resources}

Resources used to type a scrutinee that is not in WHNF \emph{must not} be used in
the case alternatives.
%
However, evaluating the scrutinee to WHNF is not sufficient to
consume all resources used by the scrutinee, since sub-expressions such
as constructor arguments will be left unevaluated and may still contain
(``parts of'') the linear resources. To \emph{fully} consume all
resources occurring in the scrutinee, all linear components of the
constructor must also be fully evaluated, as witnessed by the
$\textsc{Alt}_0$ rule above.
%
In essence, to type cases scrutinizing expressions not in WHNF, we need to
forbid the usage of the scrutinee resources while tracking that we necessarily
``finish consuming'' them, either via the case binder, or by linearly using all
linear pattern-bound variables introduced in the alternative.

@@

We introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
to encode linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
\emph{exactly once}. However, since proof irrelevant resources cannot be
discarded nor used directly, they have to be consumed \emph{indirectly} --
by $\D$-bound variables, namely, the case binder, or the linear pattern-bound variables.

>>>>

\subsubsection{Proof irrelevant resources}

When the scrutinee is not in WHNF, it is not known statically which of the
scrutinee resources end up being consumed during evaluation and which would
still be available in the branch continuation.
%
Therefore, cases with non-WHNF scrutinees must by typed such that no resource
from the scrutinee can be used directly in the branch again, or risk
consuming the same resource twice.
%
However, simply evaluating the scrutinee to WHNF does not guarantee all
resources are consumed (unless the result matches an unrestricted pattern).
In the case where the matching pattern has some linear components

is not sufficient to
consume all resources used by the scrutinee, since sub-expressions such
as constructor arguments will be left unevaluated and may still contain
(``parts of'') the linear resources. To \emph{fully} consume all
resources occurring in the scrutinee, all linear components of the
constructor must also be fully evaluated, as witnessed by the
$\textsc{Alt}_0$ rule above.
%
In essence, to type cases scrutinizing expressions not in WHNF, we need to
forbid the usage of the scrutinee resources while tracking that we necessarily
``finish consuming'' them, either via the case binder, or by linearly using all
linear pattern-bound variables introduced in the alternative.

@@

We introduce \emph{proof irrelevant} resources, written as linear resources
within square brackets $[ x{:}\s]$ and lifted to contexts as $[\D]$,
to encode linear resources that cannot be
directly used (i.e.,~to which the $\textsc{Var}$ rule is not applicable). Proof irrelevant resources
are linear resources in all other senses, meaning they must be used
exactly once. However, since proof irrelevant resources cannot be
discarded nor used directly, they have to be consumed \emph{indirectly} via
$\D$-bound variables, namely, the case binder, or the linear pattern-bound
variables.
